Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2025-03-21 paper/paper.tex
turing machine

Because the states of $D$ are $\epsilon$-closed subsets of $N.\states$, if two
states $s$ and $s'$ belong to the the same $\epsilon$-closed subset
$X : \mathbb{P}_{\epsilon}(N.\states)$ then there exists a $\epsilon$-path in
$N$ between the two, but we do not necessarily know which one. Similarly, the
data contained in the type $\Trace_{\D}}~X~true$ is the \emph{existence} of an accepting trace in
$\N$ beginning at some state in $\D$.

@@

This reification operation on functions $\StringSem \to U$ is incredibly expressive,
as it allows to sidestep our linear typing connectives and utilize the whole of nonlinear
dependent type theory to define a grammar. For example, given a
Turing machine $Tur$ one may
define a non-linear predicate $accepts : \StringSem \to U$ such that $accept~w$ is equal to $\top$ if
$T$ accepts $w$ and equal to $0$ otherwise. Then, $Reify~accepts$ is a linear
type that captures precisely the strings accepted by $Tur$. That is,
$Lang(Reify~accepts)$ is recursively enumerable --- the most general class of
languages in the Chomsky hierarchy.
\begin{theorem}[\Agda]
  For any Turing machine, we can construct a grammar in \theoryabbv
  that accepts the same language as the Turing machine.
\end{theorem}

@@

%% \subsubsection{Typed Grammars}
The usage of a simple type system to reason about regular expressions up to weak
equivalence was
introduced by Frisch and Cardelli \cite{frischCardelli}. \steven{Frisch and
  Caredilli also define a linear time greedy parser for regular expressions. We
  have not yet tried to verify a greedy algorithm, however the approach used in
  \cref{thm:determinization} would apply to their algorithm.
  \cref{thm:determinization} requires a choice function for traces through an
  NFA (i.e. a disambiguation strategy) which may be instantiated to encode greediness. } Henglein and Nielsen
\cite{henglein_regular_2011} later adapt this approach by providing a type
system for regular expressions whose semantics were in one-to-one correspondence
with the set of parse trees.
Elliott \cite{elliottSymbolicAutomaticDifferentiation2021} gives a similar
semantic interpretation of regular grammars as type-level predicates on strings. The type system of \theoryabbv extends these semantic interpretations of to a broader class of grammars and also gives a formal syntax and equational theory for the parse transformers that these prior works lack.

%% Krishnaswami and Yallop \cite{krishnaswami_typed_2019} provide a type system for
%% demonstrating that carries proves of the LL(1) condition for context-free
%% grammars. Despite being a type system over formal grammars, their interpretation
%% of types is very different from the semantics of \theoryabbv.

@@

@@

@@

Our verification has mainly focused on the verification that a parser
outputs a correct concrete syntax tree for a grammar. However, in
practice, parsers are combined with a \emph{semantic action} that
emits an \emph{abstract} syntax tree that omits superfluous syntactic
details that aren't needed in later stages of the overall program. We
can define a semantic action in \theoryabbv for a linear type $A$ with
semantic outputs in a non-linear type $X$ to be a function $\ltonl {(A
  \lto \bigoplus_{\_:X} \top)}$. That is, a semantic action is a
function that produces a semantic element of $X$ from the concrete
parses of $A$. In future work, we aim to study the question of
verifying efficient implementations of parsers with semantic actions.

>>>>

Because the states of $D$ are $\epsilon$-closed subsets of $N.\states$, if two
states $s$ and $s'$ belong to the the same $\epsilon$-closed subset
$X : \mathbb{P}_{\epsilon}(N.\states)$ then there exists a $\epsilon$-path in
$N$ between the two, but we do not necessarily know which one. Similarly, the
data contained in the type $\Trace_{\D}~X~true$ is the \emph{existence} of an accepting trace in
$\N$ beginning at some state in $\D$.

@@

This reification operation on functions $\StringSem \to U$ is incredibly expressive,
as it allows to sidestep our linear typing connectives and utilize the whole of nonlinear
dependent type theory to define a grammar. For example, given a
Turing machine $T$ one may
define a non-linear predicate $accepts : \StringSem \to U$ that encodes that $T$
halts and accepts an input string. Then, $Reify~accepts$ is a linear type that
captures precisely the string the same language as $T$. That is,
$Lang(Reify~accepts)$ is recursively enumerable --- the most general class of
languages in the Chomsky hierarchy.
\begin{construction}[\Agda]
  \label{cons:turing}
  For any Turing machine $T$, we can construct a grammar in \theoryabbv
  that accepts the same language as $T$.
\end{construction}

@@

%% \subsubsection{Typed Grammars}
Frisch and Cardelli introduced the use of a simple type system to reason about regular expressions up to weak equivalence \cite{frischCardelli}.
Henglein and Nielsen build on this type-theoretic view of grammars, pointing out that the values of the types correspond precisely to the parse trees \cite{henglein_regular_2011}.
Our design of \theoryabbv extends this view of grammars as types to a much broader class of grammars, as well as providing a syntax and equational theory for \emph{parse transformers}, showing that the terms can be viewed not just as parse trees, but as intrinsically verified parsers.

Elliott \cite{elliottSymbolicAutomaticDifferentiation2021} uses essentially the same denotational semantics as ours, interpreting regular expressions as type-valued functions on strings. Our denotational semantics of \theoryabbv extends this to a much broader class of grammars.
%% Krishnaswami and Yallop \cite{krishnaswami_typed_2019} provide a type system for
%% demonstrating that carries proves of the LL(1) condition for context-free
%% grammars. Despite being a type system over formal grammars, their interpretation
%% of types is very different from the semantics of \theoryabbv.

@@

\paragraph{Dependent Linear Types}

Our syntax for non-commutative dependent linear type theory is based on the
dependent commutative linear type theory of
\citet{krishnaswami_integrating_2015}, itself an extension of Benton and
Wadler's linear-non-linear calculus \cite{bentonMixedLinearNonlinear1995}. A
distinct feature of these systems is that linear types and non-linear types are
distinct sorts, so the linear logic $!$ operator is not a primitive, but is
instead definable. Vakar develops a dependent linear type theory where instead
the non-linear types are accessed using the $!$ modality, similar to Girard's
original approach \cite{vakar,girard_linear_1987}. Additionally, Vakar develops
a general categorical semantics for their system, whereas we have only developed
a single denotational model. It appears straightforward to adapt Vakar's general
semantics approach to a non-commutative variant that would apply to
\theoryabbv. This may have applications in finding alternative models, or
developing logical relations proofs categorically.

@@

\subsection{Future Work}

In this work we have demonstrated feasibility of \theoryabbv for the
verification of formal grammar theory and sound-by-construction parsers. In
future work, we aim to extend \theoryabbv to be a practical tool for developing
verified parsing components of larger verified software systems.

\paragraph{Verified Parser Generators}

We aim to extend our work on parsing to verify other algorithms. For instance,
our regular expression parser makes somewhat arbitrary choices in what parse to
produce when the grammar is ambiguous. In the future we aim to verify Frisch and
Cardelli's \emph{greedy} algorithm for regular expression
disambiguation. Defining this algorithm in \theoryabbv would provide soundness
by construction, but it is not obvious if the greediness property could be
verified easily as well.

We also aim to adapt the approach used for our context-free grammar parserss for
LL and LR/LALR with the aim of developing a shared library of intrinsically
verified parsing utilities that can be usefully shared across different
parsers. We would also like to investigate how high-performance verified parsers
can be implemented in \theoryabbv. This might be done by developing an efficient
compiler for \theoryabbv directly or by developing a verified compiler to an
imperative system within \theoryabbv itself.

@@

Our verification has mainly focused on the verification that a parser outputs a
correct concrete syntax tree for a grammar. However, in practice, parsers are
combined with a \emph{semantic action} that emits an \emph{abstract} syntax tree
that omits superfluous syntactic details that are unneeded in later stages of
the overall program. We can define a semantic action in \theoryabbv for a linear
type $A$ with semantic outputs in a non-linear type $X$ to be a function $\ltonl
{(A \lto \bigoplus_{\_:X} \top)}$. That is, a semantic action is a function that
produces a semantic element of $X$ from the concrete parses of $A$. In future
work, we aim to study the question of verifying efficient implementations of
parsers with semantic actions, and integrating them into larger verified
systems.
